Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-09-30 LatticeOptics/submission_rev_sci_inst/latticeoptics.tex
reviewer2 specific comments

In the following we give an in-depth technical description of the optical lattice setup we developed. We start with the laser system itself which includes the seed laser, amplification and optical setup for controlling the laser beams. In the next section we provide a detailed description of the two-stage laser power feedback system. In the last two sections we conduct a performance analysis of the short- and long-term stability of the optical lattices produced by this system.

@@

Such trap frequencies may only be attained using tightly focused, high-power beams.
In practical terms, this requires approximately \unit{20}{W} per lattice axis for a beam waist of around $80 \mu\mathrm{m}$ and the four-fold enhancement in optical depth provided by the reflection from the superpolished substrate. We achieve such high laser powers using Yb-doped fiber amplifiers (Nufern NuAmp SUB1174-34).
These amplifiers can amplify \unit{150}{mW} of seed light to up to approximately \unit{45}{W}. They use a length of doped fiber pumped with \unit{808}{nm} light from fiber coupled diode sources to amplify the \unit{1064}{nm} seed light.
The amplification is performed using a two-stage design, where the stages are controlled and powered using separate power supplies and electronics.

@@

We need to maintain the relatively low phase-noise characteristics of the VCO to prevent phase- to amplitude-noise conversion in the AOM, which partially downconverts the VCO phase noise to RIN at acoustic frequencies.
The spectrum and relative intensity noise of the PLL can be found in Fig.~\ref{fig:pll}.
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of a typical VCO, without the slow frequency drift characteristic to these devices.
Elimination of the carrier drift is worth the added phase noise, however, because any drift of the frequency affects the pointing of the laser beam passing through the AOM, which would misalign the optical lattice.

@@

Fast, low-noise photodiodes are required both to provide the monitoring arm of the feedback loop and to characterize the RIN of the system.
An ideal device would have a bandwidth up to approximately \unit{3}{MHz} (since trap frequencies in the system are \unit{1.5}{MHz}), with a noise floor below approximately \unit{-150}{dBc/Hz} over that range~\cite{Blatt2015}.
To approach these specifications, we use a custom-designed PCB that implements a simple transimpedance amplifier~\cite{Graeme1995} in a convenient physical package, providing a bandwidths in the \unit{0.1-10}{MHz} range, depending on the transimpedance gain set by a resistor.
The photodiode can be safely operated up to incident powers of approximately \unit{10}{mW}, which sets a \unit{-164}{dBc/Hz} shot noise limit on the RIN that can be detected.
In practice, we typically operate the photodiodes with $\approx$\unit{1-5}{mW} of incident power.
A photodiode with a bandwidth (BW) of \unit{6}{MHz} is used to characterize the system, while photodiodes with a bandwidth of \unit{330}{kHz} (and correspondingly higher gain) are used to perform the feedback.

@@

The connection between the actuation and monitoring mechanisms of the feedback system is the loop filter, which compares the \emph{set-point} signal, $V_\mathrm{SP}(t)$, supplied by the experimental control system and the photodiode signals $V_\mathrm{PD}(t)$, to adjust the actuation accordingly.
We use a simple, purely integrating design in three stages, as seen in Fig.~\ref{fig:circuits}(c), based on a low-noise op-amp (Texas Instruments OPA211).
The first stage computes and amplifies an error signal, $e(t)=V_\mathrm{SP}(t)-V_\mathrm{PD}(t)$, which is integrated by the second stage.
The integral gain is set using a low-drift trim-pot.
This is the case because within relevant frequency ranges, the AOM can be coarsely approximated by a pure time delay process, given by the acoustic wave transit time of $\approx$\unit{1}{\mu{}s} across the active region of the AOM crystal.
In this scenario, the proportional term must be small to avoid instability at the $\pi$ phase shift point, and a pure integral controller is optimal~\cite{Skogestad2003}.
For debugging purposes, the loop filter features buffered photodiode monitor and error signal monitor outputs.

@@

Several applications require the positions of lattice sites to not change by more than a small fraction ($\approx 10\%$) between experimental shots.
In our setup this is ensured by careful design of the optical assembly, where the phase of the retro-reflected lattice is set by the retro-reflecting mirror, which is placed approximately \unit{5}{cm} from the atoms~\cite{Huber2014}.
This phase determines the location of the sites of the optical lattice.
We have characterized the phase stability of the optical lattice by recording a long sequence of densely populated images with our imaging camera. We found that the phase varies by much less than one lattice site over the course of hours, as seen in Fig.~\ref{fig:averaged_mott}.

@@

\section{Long-term stability of the lattice structure}
A benefit of site-resolved imaging is that every measurement of the atomic distribution is also a direct measurement of the underlying lattice structure (except for the short-term and small-scale displacement discussed in the last section).
As discussed in the supplementary material of~\cite{Parsons2016, Greif2016}, the Fourier transform of an atomic distribution contains distinct peaks corresponding to the lattice structure, and these peaks can be used to extract the relative angle and the spacing of the lattices.
For this reason, we recorded a long-term measurement spanning almost one year to learn how these parameters change over time, see Fig.~\ref{fig:long_term_stability}.
Several interesting features can be seen from the observed behavior. In the first few months the lattice spacing changed by around 10\%. Experimentally, this drift was accompanied by the necessity of refocussing the imaging system. The sudden jump in September 2016 was caused by a simplification of the imaging system. Afterwards, the lattice spacing seemed more stable (fractional changes of $\leq 0.005$). Unlike the lattice spacing, the angle is observed to be mostly unaffected and remains very stable (for example, an RMS of 0.06$^\circ$ in Spring 2017).

>>>>

In the following we give an in-depth technical description of the optical lattice setup we developed. We start with the laser system itself which includes the seed laser, amplification and optical setup for controlling the laser beams. In the next section we provide a detailed description of the two-stage laser power feedback system. In the last two sections we conduct a performance analysis of the short- and long-term stability of the optical lattices produced by this system. 
Naturally, the laser system presented here is tailored to suit the experiment's specific lattice geometry~\cite{Huber2014} and exact requirements. 
However, we expect that the technical upgrades, design practices and engineering choices will be widely applicable to experimentalists working with performant optical lattices.

@@

Such trap frequencies may only be attained using tightly focused, high-power beams.
In practical terms, this requires approximately \unit{20}{W} per lattice axis for a beam waist of around $80 \mu\mathrm{m}$ and the four-fold enhancement in optical depth provided by the reflection from the superpolished substrate~\cite{Huber2014}. We achieve such high laser powers using Yb-doped fiber amplifiers (Nufern NuAmp SUB1174-34).
These amplifiers can amplify \unit{150}{mW} of seed light to up to approximately \unit{45}{W}. They use a length of doped fiber pumped with \unit{808}{nm} light from fiber coupled diode sources to amplify the \unit{1064}{nm} seed light.
The amplification is performed using a two-stage design, where the stages are controlled and powered using separate power supplies and electronics.

@@

%We need to maintain the relatively low phase-noise characteristics of the VCO to prevent phase- to amplitude-noise conversion in the AOM, which partially downconverts the VCO phase noise to RIN at acoustic frequencies.
This device allows us to drive the AOM with a single, clean tone.
The spectrum and relative intensity noise of the PLL can be found in Fig.~\ref{fig:pll}.
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of a typical VCO, without the slow frequency drift characteristic to these devices.
Elimination of the carrier drift is worth the added phase noise, however, because any drift of the frequency affects the pointing of the laser beam passing through the AOM, which would misalign the optical lattice.

@@

Fast, low-noise photodiodes are required both to provide the monitoring arm of the feedback loop and to characterize the RIN of the system.
An ideal device would have a bandwidth up to approximately \unit{3}{MHz} (since trap frequencies in the system are \unit{1.5}{MHz}), with a noise floor below approximately \unit{-150}{dBc/Hz} over that range~\cite{Blatt2015}.
To approach these specifications, we use a custom-designed PCB that implements a simple transimpedance amplifier~\cite{Graeme1995} in a convenient physical package, providing bandwidths in the \unit{0.1-10}{MHz} range, depending on the transimpedance gain set by a resistor.
The photodiode can be safely operated up to incident powers of approximately \unit{10}{mW}, which sets a \unit{-164}{dBc/Hz} shot noise limit on the RIN that can be detected.
In practice, we typically operate the photodiodes with $\approx$\unit{1-5}{mW} of incident power.
A photodiode with a bandwidth (BW) of \unit{6}{MHz} is used to characterize the system, while photodiodes with a bandwidth of \unit{330}{kHz} (and correspondingly higher gain) are used to perform the feedback.

@@

The connection between the actuation and monitoring mechanisms of the feedback system is the loop filter, which compares the \emph{set-point} signal, $V_\mathrm{SP}(t)$, supplied by the experimental control system and the photodiode signals $V_\mathrm{PD}(t)$, to adjust the actuation accordingly.
We use a simple, primarily integrating design in three stages, as seen in Fig.~\ref{fig:circuits}(c), based on a low-noise op-amp (Texas Instruments OPA211).
The first stage computes and amplifies an error signal, $e(t)=V_\mathrm{SP}(t)-V_\mathrm{PD}(t)$, which is integrated by the second stage.
The integral gain is set using a low-drift trim-pot.
This is the case because within relevant frequency ranges, the AOM can be coarsely approximated by a pure time delay process, given by the acoustic wave transit time of $\approx$\unit{1}{\mu{}s} across the active region of the AOM crystal.
In this scenario, the proportional term must be small to avoid instability at the $\pi$ phase shift point, and a pure integral controller is optimal~\cite{Skogestad2003}.
For debugging purposes, the loop filter features buffered photodiode monitor and error signal monitor outputs.

@@

Several applications require the positions of lattice sites to not change by more than a small fraction ($\approx 10\%$ of the lattice constant) between experimental shots.
In our setup this is ensured by careful design of the optical assembly, where the phase of the retro-reflected lattice is set by the retro-reflecting mirror, which is placed approximately \unit{5}{cm} from the atoms~\cite{Huber2014}.
This phase determines the location of the sites of the optical lattice.
We have characterized the phase stability of the optical lattice by recording a long sequence of densely populated images with our imaging camera. We found that the phase varies by much less than one lattice site over the course of hours, as seen in Fig.~\ref{fig:averaged_mott}.

@@

\section{Long-term stability of the lattice structure}
A benefit of site-resolved imaging is that every measurement of the atomic distribution is also a direct measurement of the underlying lattice structure (except for the short-term and small-scale displacement discussed in the last section).
As discussed in the supplementary material of~\cite{Parsons2016, Greif2016}, the Fourier transform of an atomic distribution contains distinct peaks corresponding to the lattice structure, and these peaks can be used to extract the relative angle and the spacing of the lattices.
For this reason, we recorded a long-term measurement spanning almost one year to learn how these parameters change over time, see Fig.~\ref{fig:long_term_stability}.
Several interesting features can be seen from the observed behavior. In the first few months the apparent lattice spacing (as imaged through the microscope) changed by around 10\%. Experimentally, this drift was accompanied by the necessity of refocusing the imaging system. The sudden jump in September 2016 was caused by a simplification of the imaging system. Afterwards, the lattice spacing seemed more stable (fractional changes of $\leq 0.005$ and limited by the imaging system magnification drift). Unlike the lattice spacing, the angle is observed to be mostly unaffected and remains very stable (for example, an RMS of 0.06$^\circ$ in Spring 2017).
